An item response theory analysis of the matrix reasoning item bank (MaRs-IB)

### 2022-01-19 manuscript/main.tex
push edits

To ensure data quality, the data from multiple participants who completed the experiment were excluded prior to analysis for one or more of the following reasons: for rapid guessing -- defined here as responding in less than 3 s -- on 4 or more trials N=70); for failing to respond on four or more trials (N=8); for minimizing or otherwise shrinking the browser window beyond what was required to view the stimuli (N=7). In total, 83 of 1584 (5.24\%) participants were excluded, leaving the data from N=1501 participants for analysis. Among these participants, each item (64 in total) was sampled approximately 375 times (sd = 12) and each item variant (384 in total) was sampled approximately 62 times (sd = 7).

@@

After providing consent, participants completed the experiment which was divided into three section. In the first, participants completed three short surveys: the 10-item need for cognition survey \citep{chiesi2018applying}, the 8-item PROMIS Cognitive Function-Short Forms (8a) \citep{iverson2021normative}, and the 8-item subject numeracy scale \citep{fagerlin2007measuring}. These measures were included as part of exploratory analyses that were not the central focus of this study and will not be reported in the main text (reported in Supplement). Next, participants completed one of the three 12-item MaRs-IRB short-forms described above. Besides for the items, the administration procedure of the short-forms was identical to that in the Calibration study. Finally, participants completed the 9-item abbreviated Raven's progressive matrices (form A) \citep{bilker2012development}. For consistency's sake, the administration procedure was matched to the MaRs-IB.

The data from a total of N=47 participants were excluded prior to analysis. Of these, N=17 participants were excluded for having not completed the entire experiment. Separately, the data from N=21 participants were excluded for failing one or more attention checks (infrequency items) embedded in the self-report measures \citep{zorowitz2021inattentive}. The data from N=13 participants were excluded for image loading problems on either the MaRs-IB or RPM tasks. Finally, N=6 participants were excluded for missing or rapid guessing on 4 or more trials during the MaRs-IB or RPM tasks. Note these are not mutually exclusive.

The majority of participants in the validation sample identified as women (men: N=138; women: N=147; non-binary or other: N=13; rather not say: N=2). Participants were 35.0 years old (13.2 sd) on average. The sample was relatively well-educated with the majority having a Bachelor's degree (N=121) or Master's degree or higher (N=45); fewer participants endorsed only completing some college (N=100) or only a high school degree or lower (N=33).

\subsection{Results}

The distribution of scores on the three MaRs-IB short-forms are summarized in Table \ref{table:2}. Across the short-forms, participants responded correctly on 8.0 of 12 items (sd = 2.5, IQR = 6 - 10). The distribution of scores were skewed towards to higher end, as would be expected given their respective test characteristic curves. A 1-way ANOVA did not reject the null hypothesis that the observed scores across test forms were equal (F(2,97) = 0.253, p = 0.777). Corroborating the predictions from the test assembly, the reliability of each short form was satisfactory (Table \ref{table:2}).

The distribution of scores on the RPM short-form is similarly summarized in Table \ref{table:2}. On average, participants responded correctly on 4.5 of 9 items (sd = 2.0, IQR = 3 - 9). Scores on the RPM were significantly lower on the RPM than the MaRs-IB ($t = 25.598, p < 1e^{-6}$), indicating that the items on the RPM short-form are more difficult on average than those selected for the MaRs-IB short-forms. The reliability of the measure was similarly good as the MaRs-IB (Table \ref{table:2}).

@@

\begin{table}
    \centering
    \begin{tabular}{c|c|c|c}
    Measure & Mean (sd) & IQR & Reliability \\
    \hline
    MaRs-IB SFa & 7.9 (2.4) & 7 - 10 & 0.761 \\
    MaRs-IB SFb & 8.2 (2.6) & 7 - 10 & 0.810 \\
    MaRs-IB SFc & 7.9 (2.7) & 6 - 10 & 0.829 \\
    RPM SFa & 4.5 (2.0) & 3 - 9 & 0.791 \\
    \end{tabular}
    \caption{Caption}
    \label{table:2}
\end{table}

>>>>

To ensure data quality, the data from multiple participants who completed the experiment were excluded prior to analysis for one or more of the following reasons: for rapid guessing -- defined here as responding in less than 3 s -- on 4 or more trials (N=70); for failing to respond on four or more trials (N=8); for minimizing or otherwise shrinking the browser window beyond what was required to view the stimuli (N=7). In total, 83 of 1584 (5.24\%) participants were excluded, leaving the data from N=1501 participants for analysis. Among these participants, each item (64 in total) was sampled approximately 375 times (sd = 12) and each item variant (384 in total) was sampled approximately 62 times (sd = 7).

@@

After providing consent, participants completed the experiment which was divided into three section. In the first, participants completed three short surveys: the 10-item need for cognition survey \citep{chiesi2018applying}, the 8-item PROMIS Cognitive Function-Short Forms (8a) \citep{iverson2021normative}, and the 8-item subject numeracy scale \citep{fagerlin2007measuring}. These measures were included as part of planned exploratory analyses and will not be reported in the main text (all correlations between these self-report measures and performance on the matrix reasoning tasks are reported in Supplement). Next, participants completed one of the three 12-item MaRs-IRB short-forms described above. The administration procedure of the short-forms was identical to that in the first study. Finally, participants completed the 9-item abbreviated Raven's progressive matrices (form A) \citep{bilker2012development}. To minimize bias, the administration procedure for the progressive matrices was matched to that for the MaRs-IB items.

To ensure data quality, the data from multiple participants were excluded prior to analysis for one or more of the following reasons: for failing to complete all three sections of the experiment (N=17); for failing one or more attention checks \citep{zorowitz2021inattentive} in the self-report measures (N=21); for experiencing technical difficulties during the loading of the stimuli (N=11); for rapid guessing on 4 or more trials (N=3); or for failing to respond on four or more trials (N=1). In total, 47 of 347 (13.54\%) participants were excluded, leaving the data from N=300 participants for analysis.

The majority of participants identified as women (men: N=138; women: N=147; non-binary or other: N=13; rather not say: N=2). Participants were 35.0 years old (13.2 sd) on average. The sample was relatively well-educated, with the majority having completed a Bachelor's degree (N=121) or Master's degree or higher (N=45). By comparison, fewer participants endorsed having completed only some college (N=100), only a high school degree (N=33), or preferring not to say (N=1).

The distribution of scores on the three MaRs-IB short-forms are summarized in Table \ref{table:2}. Consistent with model predictions, participants on average responded correctly on 8.0 of 12 items (sd = 2.5, IQR = 6 - 10). Similarly, the distribution of total scores were right-shifted \ref{fig:fig03} as would be expected given the nominal guessing rate of 25\%. A one-way ANOVA comparing the total scores across the three short-forms did not reject the null hypothesis (F(2,297) = 0.253, p = 0.777). Finally, the marginal score reliability for each short form was satisfactory as predicted (Table \ref{table:2}). In summary, the distribution of scores across the MaRs-IB short forms were consistent with the predictions based on the test assembly, indicating that the psychometric properties of these items were well calibrated in the previous study.

The distribution of scores on the abbreviated Raven's progressive matrices is also summarized in Table \ref{table:2}. On average, participants responded correctly on 4.5 of 9 items (sd = 2.0, IQR = 3 - 9). Scores on the RPM were significantly lower on the RPM than the MaRs-IB ($t = 25.598, p < 1e^{-6}$), indicating either that the RPM items are more difficult or less susceptible to guessing (consistent with these items having 6 or 8 distractors). The marginal score reliability for the RPM was similar to that of the MaRs-IB forms (Table \ref{table:2}).

\begin{figure}
\centering
\includegraphics[width=1.0\textwidth]{figures/fig03.png}
\caption{\label{fig:fig03}This is a figure caption.}
\end{figure}

\subsection{Convergent validity}

@@

\begin{table}
    \centering
    \begin{tabular}{c|c|c|c}
    Measure & Mean (sd) & IQR & Reliability \\
    \hline
    MaRs-IB SF-1 & 7.9 (2.4) & 7 - 10 & 0.761 \\
    MaRs-IB SF-2 & 8.2 (2.6) & 7 - 10 & 0.810 \\
    MaRs-IB SF-3 & 7.9 (2.7) & 6 - 10 & 0.829 \\
    RPM SF-A & 4.5 (2.0) & 3 - 9 & 0.791 \\
    \end{tabular}
    \caption{Caption}
    \label{table:2}
\end{table}
